---
title: "DEiR01"
author: "Robert A. Stevens"
date: "July 1, 2016"
output: html_document
---

```{r, comment=FALSE}
library(deSolve)
library(scatterplot3d)
```

# Differential Equations in R

by Karline Soetaert & Thomas Petzoldt

Tutorial useR conference, June 24, 2014

http://desolve.r-forge.r-project.org/user2014/

## Introduction

### Outline

- How to specify a model

- An overview of solver functions

- Plotting, scenario comparison

- Fitting models to data

- Forcing functions and events

- Partial differential equations with ReacTran

- Speeding up

## Installing

Download R from the CRAN website: http://cran.r-project.org/

Install a suitable editor, e.g. RStudio

Packages can be installed from within R or Rstudio

or via the command line

    install.packages("deSolve", dependencies = TRUE)

### Several packages deal with differential equations

- deSolve: main integration package

- rootSolve: steady-state solver

- bvpSolve: boundary value problem solvers

- deTestSet: ODE and DAE test set + additional solvers

- ReacTran: partial differential equations

- simecol: interactive environment for implementing models

All packages have at least one author in common → consistent interface.

More, see CRAN Task View: Differential Equations

## Getting help

```
?deSolve
```

- opens the main help page containing
    + a short explanation
    + a link to the main manual (vignette) “Solving Initial Value Differential Equations in R”
    + links to additional manuals, papers and online resources
    + references
    + a first example

- all our packages have such a ?<packagename> help file.

```
example(deSolve)
```

```{r, comment=FALSE}
# Chaos in the atmosphere
Lorenz <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    dX <-  a*X + Y*Z
    dY <-  b*(Y - Z)
    dZ <- -X*Y + c*Y - Z
    list(c(dX, dY, dZ))
  })
}
parameters <- c(a = -8/3, b = -10, c = 28)
state      <- c(X = 1, Y = 1, Z = 1)
times      <- seq(0, 100, by = 0.01)
out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
plot(out)
# add a 3D figure if package scatterplot3D is available
scatterplot3d(out[, -1], type = "l")
```
