---
title: "DEiR02"
author: "Robert A. Stevens"
date: "July 1, 2016"
output: html_document
---

```{r, comment=FALSE}
library(deSolve)
library(scatterplot3d)
```

# Differential Equations in R 

by Karline Soetaert & Thomas Petzoldt

## Introductory example

### Let’s begin...

Model specification

### Logistic growth 

Differential equation

dN/dt = r·N·(1 − N/K)

Analytical solution

Nt = K\*N0\*exp(r\*t)/[K + N0\*(exp(r\*t) − 1)]

R implementation

```{r, comment=FALSE}
logistic <- function(t, r, K, N0) {
  K*N0*exp(r*t)/(K + N0*(exp(r*t) - 1))
}
plot(0:100, logistic(t = 0:100, r = 0.1, K = 10, N0 = 0.1))
```

### Numerical simulation in R

Why numerical solutions?

- Not all systems have an analytical solution,

- Numerical solutions allow discrete forcings, events, ...

Why R?

- If standard tool for statistics, why Prog$$$ for dynamic simulations?

- The community and the packages → useR!2014

Numerical solution of the logistic equation

```{r, comment=FALSE}
model <- function (time, y, parms) {
  with(as.list(c(y, parms)), {
  	# Differential equation "similar to formula on paper"
    dN <- r*N*(1 - N/K) 
	list(dN)})
}
y     <- c(N = 0.1)
parms <- c(r = 0.1, K = 10)
times <- seq(0, 100, 1)
# Numerical methods provided by the deSolve package
out <- ode(y, times, model, parms)
plot(out)
```

### Inspecting output

Print to screen 

```{r, comment=FALSE}
head(out, n = 4)
```

Summary

```{r, comment=FALSE}
summary(out)
```

Plotting

```{r, comment=FALSE}
plot(out, main = "logistic growth", lwd = 2)
```

Diagnostics

```{r, comment=FALSE}
diagnostics(out)
```

## Coupled equations

### Coupled ODEs: the rigidODE problem

Problem

- Euler equations of a rigid body without external forces.

- Three dependent variables (y1, y2, y3), the coordinates of the rotation vector,

- I1, I2, I3 are the principal moments of inertia

### Coupled ODEs: the rigidODE equations 

Differential equation

y1′ = (I2−I3)/I1·y2y3 

y2′ = (I3−I1)/I2·y1y3 

y3′ = (I1−I2)/I3·y1y2

Parameters

I1 = 0.5

I2 = 2

I3 = 3

Initial conditions

y1(0) = 1

y2(0) = 0

y3(0) = 0.9

### Coupled ODEs: the rigidODE problem 

R implementation

```{r, comment=FALSE}
rigidode <- function(t, y, parms) {
	dy1 <- -2.00*y[2]*y[3] 
	dy2 <-  1.25*y[1]*y[3] 
	dy3 <- -0.50*y[1]*y[2] 
	list(c(dy1, dy2, dy3))
}
yini  <- c(y1 = 1, y2 = 0, y3 = 0.9)
times <- seq(from = 0, to = 20, by = 0.01)
out   <- ode(times = times, y = yini, func = rigidode, parms = NULL)
head(out, n = 3)
```

### Coupled ODEs: plotting the rigidODE problem

```{r, comment=FALSE}
plot(out)
par(mar = c(0, 0, 0, 0)) 
scatterplot3d(out[, -1])
```

### Exercise

The Rossler equations

y1′ = −y2 − y3

y2′ = y1 + a·y2

y3′ = b + y3·(y1 − c)

Initial Conditions

y1 = 1

y2 = 1

y3 = 1 

Parameters

a = 0.2

b = 0.2

c = 5

Tasks

- Solve the ODEs on the interval [0, 100]

- Produce a 3-D phase-plane plot

- Use file examples/rigidODE.R.txt as a template

```{r, comment=FALSE}
Rossler <- function(t, y, parms) {
  # Parameters (pass into function?)
  a <- 0.2
  b <- 0.2
  c <- 5
  #
  dy1 <- -y[2] - y[3]
  dy2 <-  y[1] + a*y[2]
  dy3 <-  b + y[3]*(y[1] - c)
  list(c(dy1, dy2, dy3))
}
yini  <- c(y1 = 1, y2 = 1, y3 = 1)
times <- seq(from = 0, to = 20, by = 0.01)
out   <- ode(times = times, y = yini, func = Rossler, parms = NULL)
head(out, n = 3)

#plotting the solutions
plot(out)
par(mar = c(0, 0, 0, 0)) 
scatterplot3d(out[, -1])
```

[3] E. Hairer, S. P. Norsett, and G Wanner. Solving Ordinary Differential Equations I: Nonstiff roblems. Second Revised Edition. Springer-Verlag, Heidelberg, 2009

[13] O.E. Rossler. An equation for continous chaos. Physics Letters A, 57 (5):397-398, 1976.
