---
title: "INTRODUCTION TO SET THEORY AND SETS WITH R"
author: "Robert A. Stevens"
date: "July 16, 2017"
output: html_document
---

http://www.aaronschlegel.com/introduction-set-theory-r/

Part 1 of 5 in the series Set Theory

1. Introduction to Set Theory and Sets with R

2. Set Operations Unions and Intersections in R

3. Set Theory Arbitrary Union and Intersection Operations with R

4. Algebra of Sets in R

5. Set Theory Ordered Pairs and Cartesian Product with R

Sets define a 'collection' of objects, or things typically referred to as 'elements' or 'members.' The concept of sets arises naturally when dealing with any collection of objects, whether it be a group of numbers or anything else. Conceptually, the following examples can be defined as a 'set':

- {1, 2, 3, 4}

- {Red, Green, Blue}

- {Cat, Dog}

The first example is the set of the first four natural numbers. The second defines a set of the primary colors while the third example denotes a set of common household pets.

Since its development beginning in the 1870s with Georg Cantor and Richard Dedekind, set theory has become a foundational system of mathematics, and therefore its concepts constantly arise in the study of mathematics and is also an area of active research today.

This post will introduce some of the basic concepts of set theory, specifically the Zermelo-Fraenkel axiomatic system (more on that later), with R code to demonstrate these concepts.

## Set Notation

Sets can be defined with lowercase, uppercase, script or Greek letters (in addition to subscripts and the like). Using several types of letters helps when dealing with hierarchies. Before diving into set theory, it's best to define the common notation employed. One benefit of set theory being ubiquitousness in mathematics is learning its notation also helps in the understanding of other mathematical concepts.

- $\forall$ x - for every set x

- $\exists$ x - there exists such a set x that

- $\neg$ - not

- $\wedge$ - and

- $\vee$ - or (one or the other or both)

- $\Rightarrow$- implies

- $\Leftrightarrow$ - iff, if and only if.

- **A** $\cup$ **B** - union of sets **A** and **B**

- **A** $\cap$ **B** - intersection of sets **A** and **B**

- x $\in$ **A** - x is an element of a set **A**

- x $\notin$ **A** - x is not an element of a set **A**

The $\Rightarrow$ notation for implies can be thought of like an if statement in that it denotes the relation 'if a, then b.'

## Set Membership

Set membership is written similar to:

x $\in$ **A**

Which can be stated as 'xx is an element of the set **A**.' If xx is not a member of **A**, we write:

x $\notin$ **A**

The symbol $\in$ to denote set membership originated with Giuseppe Peano (Enderton, pp. 26).

Which is read 'x is not an element of the set **A**.' We can write an R function to implement the concept of set membership. Note there already exists a function is.element() in base R that it is recommended for practical applications.

```{r, comment=NA}
iselement <- function(x, A) {
  if(x %in% A) {
    return(TRUE)
  }
  return(FALSE)
}

my.iselement <- function(x, A) {
  return(x %in% A)
}
```

Let's use our simple function to test if there exists some members in the set, **A** = {3, 5, 7, 11}.

```{r, comment=NA}
A <- c(3, 5, 7, 11)
eles <- c(3, 5, 9, 10, (5 + 6))

for(i in 1:length(eles)) {
  print(iselement(i, A))
}

# my version
for(i in 1:length(eles)) {
  print(my.iselement(i, A))
}

# base R version
for(i in 1:length(eles)) {
  print(is.element(i, A))
}
```

Set membership leads into one of the first axioms of set theory under the Zermelo-Fraenkel system, the Principle of Extensionality.

## Principle of Extensionality

The principle of extensionality states if two sets have the same members, they are equal. The formal definition of the principle of extensionality can be stated more concisely using the notation given above:

$\forall$ **A** $\forall$ **B** ($\forall$ x (x $\in$ **A** $\Leftrightarrow$ x $\in$ **B**) $\Rightarrow$ **A** = **B**)

Stated less concisely but still using set notation:

If two sets A and B are such that for every element (member) x:

x $\in$ **A** iff x $\in$ **B**

Then **A** = **B**.

We can express this axiom through an R function to test for set equality. Base R also has a function setequal() that performs the same operation.

```{r, comment=NA}
isequalset <- function(a, b) {
  a <- unique(a)
  b <- unique(b)
  an <- length(a)
  
  if(an != length(b)) {
    return(FALSE)
  }

  for(i in 1:an) {
    if(!(a[i]) %in% b) {
      return(FALSE)
    }
  }

  return(TRUE)
}

my.isequalset <- function(a, b) {
  return(all((a %in% b)) & all((b %in% a)))
}
```

We can now put the principle of extensionality in action with our R function!

```{r, comment=NA}
# original set A to compare
A <- c(3, 5, 7, 11)

# define some sets to test for equality
B <- c(5, 7, 11, 3)
C <- c(3, 4, 6, 5)
D <- c(3, 5, 7, 11, 13)
E <- c(11, 7, 5, 3)
G <- c(3, 5, 5, 7, 7, 11)

# collect sets into a list to iterate
sets <- list(B, C, D, E, G)

# using the isequalset() function, print the results of the equality tests.
for(i in sets) {
  print(isequalset(i, A))
}

# my version
for(i in sets) {
  print(my.isequalset(i, A))
}

# base R version
for(i in sets) {
  print(setequal(i, A))
}
```

## Empty Sets and Singletons

So far we have only investigated sets with two or more members. The empty set, denoted $\varnothing$, is defined as a set containing no elements and occurs surprisingly frequently in set-theoretic operations despite is seemingly straightforward and simple nature.

The empty set axiom, states the existence of an empty set concisely:

$\exists$ **B** $\forall$ x, x $\notin$ **B**

Which can also be stated as 'there is a set having no members.'

A set $\varnothing$ can be formed whose only member is $\varnothing$. It is important to note $\varnothing$ $\neq$ $\varnothing$ because $\varnothing$ $\in$ $\varnothing$ but $\varnothing$ $\notin$ $\varnothing$. One can conceptually think of $\varnothing$ as a container with nothing in it.

A singleton is a set with exactly one element, denoted typically by *a*. A nonempty set is, therefore, a set with one or more element. Thus a singleton is also nonempty. We can define another quick function to test if a given set is empty, a singleton or a nonempty set.

```{r, comment=NA}
typeofset <- function(a) {
  if(length(a) == 0) {
    return('empty set')
  } else if(length(a) == 1) {
    return('singleton')
  } else if(length(a) > 1) {
    return('nonempty set')
  } else {
    stop('could not determine type of set')
  }
}

A <- c()
B <- c(0)
C <- c(1, 2)
D <- list(c())

set_types <- list(A, B, C, D)

for(i in set_types) {
  print(typeofset(i))
}
```

Note **D** is defined as a singleton because the set contains one element, the empty set $\varnothing$.

## Summary

This post introduced some of the basic concepts of axiomatic set theory using the Zermelo-Fraenkel axioms by exploring the idea of set, set membership and some particular cases of sets such as the empty set and singletons. Set notation that will be used throughout not just set-theoretic applications but throughout mathematics was also introduced.

## References

Barile, Margherita. "Singleton Set." From MathWorld-A Wolfram Web Resource, created by Eric W. Weisstein. http://mathworld.wolfram.com/SingletonSet.html

Enderton, H. (1977). *Elements of set theory* (1st ed.). New York: Academic Press.

Weisstein, Eric W. "Empty Set." From MathWorld-A Wolfram Web Resource. http://mathworld.wolfram.com/EmptySet.html

Weisstein, Eric W. "Nonempty Set." From MathWorld-A Wolfram Web Resource. http://mathworld.wolfram.com/NonemptySet.html
