---
title: "SET THEORY ARBITRARY UNION AND INTERSECTION OPERATIONS WITH R"
author: "Robert A. Stevens"
date: "August 3, 2017"
output: html_document
---

http://www.aaronschlegel.com/set-theory-arbitrary-unions-intersections-r/

Part 3 of 5 in the series Set Theory

1. Introduction to Set Theory and Sets with R

2. Set Operations Unions and Intersections in R

3. Set Theory Arbitrary Union and Intersection Operations with R

4. Algebra of Sets in R

5. Set Theory Ordered Pairs and Cartesian Product with R

The union and intersection set operations were introduced in a previous post using two sets, **a** and **b**. These set operations can be generalized to accept any number of sets.

## Arbitrary Set Unions Operation

Consider a set of infinitely many sets:

**A** = {$b_0$, $b_1$, $b_2$, ...}

It would be very tedious and unnecessary to repeat the union statement repeatedly for any non-trivial amount of sets, for example, the first few unions would be written as:

$b_0$ $\cup$ $b_1$ $\cup$ $b_2$ $\cup$ $b_3$ $\cup$ $b_4$ $\cup$ $b_5$

Thus a more general operation for performing unions is needed. This operation is denoted by the $\bigcup$ symbol. For example, the set **A** above and the desired unions of the member sets can be generalized to the following using the new notation:

$\bigcup$ **A** = $\bigcup_i$ $b_i$

We can then state the following definition: For a set **A**, the union $\bigcup$ **A** of **A** is defined by:

$\bigcup$ **A** = {x | ($\exists$ b $\in$ **A**) x $\in$ b}

For example, consider the three sets:

a = {2, 4, 6}

b = {3, 5, 7}

c = {2, 3, 8}

The union of the three sets is written as:

$\bigcup$ {{2, 4, 6}, {3, 5, 7}, {2, 3, 8}} = {2, 3, 4, 5, 6, 7, 8}

Recalling our union axiom from a previous post, the union axiom states for two sets **A** and **B**, there is a set whose members consist entirely of those belonging to sets **A** or **B**, or both. More formally, the union axiom is stated as:

$\forall$ a $\forall$ b $\exists$ **B** $\forall$ x (x $\in$ **B** $\Leftrightarrow$ x $\in$ a $\vee$ x $\in$ b)

As we are now dealing with an arbitrary amount of sets, we need an updated version of the union axiom to account for the change.

Restating the union axiom:

For any set **A**, there exists a set **B** whose members are the same elements of the elements of **A**. Stated more formally:

$\forall$ x [x $\in$ **B** $\Leftrightarrow$ ($\exists$ b $\in$ **A**) x $\in$ b]

The definition of $\bigcup$ **A** can be stated as:

x $\in$ $\bigcup$ **A** $\Leftrightarrow$ ($\exists$ b $\in$ **A**) x $\in$ b

For example, we can demonstrate the updated axiom with the union of four sets {a, b, c, d}:

$\bigcup$ {a, b, c, d} = {($\exists$ **B** $\in$ **A**) x $\in$ {a, b, c, d}}

$\bigcup$ {a, b, c, d} = a $\cup$ b $\cup$ c $\cup$ d

We can implement the set operation for an arbitrary amount of sets by expanding upon the function we wrote previously.

Update: Hoelk graciously provided much easier to read and faster routines for performing unions and intersection on any number of sets in the comments below.

```{r, comment=NA}
set.unions <- function(a, b, ...) {
  u <- a
  for(i in 1:length(b)) {
    if(!(b[i] %in% u)) {
      u <- append(u, b[i])
    }
  }

  s <- list(...)

  for(i in s) {
    for(j in i) {
      if(!(j %in% u)) {
        u <- append(u, j)
      }
    }
  }

  return(u)
}

my.set.unions <- function(a, b, ...) {
  u <- c(a, b, ...) 
  return(sort(unique(u)))
}
```

Perform the set union operation of four sets:

a = {1, 2, 3}

b = {3, 4, 5}

c = {1, 4, 6}

d = {2, 5, 7}

```{r, comment=NA}
a <- c(1, 2, 3)
b <- c(3, 4, 5)
c <- c(1, 4, 6)
d <- c(2, 5, 7)

set.unions(a, b, c, d)

# my version
my.set.unions(a, b, c, d)

# base R version
Reduce(union, list(a, b, c, d))
```

## Intersections of an Arbitrary Number of Sets

The intersection set operation can also be generalized to any number of sets. Consider the previous set containing an infinite number of sets.

**A** = {$b_0$, $b_1$, $b_2$, ...}

As before, writing out all the intersections would be tedious and not elegant. The intersection can instead be written as:

$\bigcap$ **A** = $\bigcap_i$ $b_i$

As before in our previous example of set intersections, there is no need for a separate axiom for intersections, unlike unions. Instead, we can state the following theorem, for a nonempty set **A**, a set **B** exists that such for any element x:

x $\in$ **B** $\Leftrightarrow$ x $\in$ $\forall$ **A**

Consider the following four sets:

a = {1, 2, 3, 5}

b = {1, 3, 5}

c = {1, 4, 5, 3}

d = {2, 5, 1, 3}

The intersection of the sets is written as:

$\bigcap$ {{1, 2, 3, 5}, {1, 3, 5}, {1, 4, 5, 3}, {2, 5, 1, 3}} 

= {1, 2, 3, 5} $\cap$ {1, 3, 5} $\cap$ {1, 4, 5, 3} $\cap$ {2, 5, 1, 3} 

= {1, 3, 5} 

We can write another function to implement the set intersection operation given any number of sets.

```{r, comment=NA}
set.intersections <- function(a, b, ...) {
  intersect <- vector()
  
  for(i in a) {
    if (i %in% b) {
      intersect <- append(intersect, i)
    }
  }

  s <- list(...)

  for(i in s) {
    for(j in i) {
      if (j %in% intersect) {
        intersect <- append(intersect, j)
      }
    }
  }

  intersect <- unique(intersect)
  return(intersect)
}

my.set.intersections <- function(a, b, ...) {
  inter <- intersect(a, b)
  s <- list(...)

  for(i in s) {
    inter <- intersect(inter, i)
  }

  return(sort(inter))
}
```

Perform set intersections of the four sets specified earlier.

```{r, comment=NA}
a <- c(1, 2, 3, 5)
b <- c(1, 3, 5)
c <- c(1, 4, 5, 3)
d <- c(2, 5, 1, 3)

set.intersections(a, b, c, d)

# my version
my.set.intersections(a, b, c, d)

# base R version
Reduce(intersect, list(a, b, c, d))
```

## References

Enderton, H. (1977). *Elements of Set Theory* (1st ed.). New York: Academic Press.
